---
title: "Applying CSIDE to Spatial Transcriptomics Data: Differential Expression with Multiple Experimental Replicates"
author: "Dylan Cable"
date: "December 21st, 2021"
output:
  html_document:
    keep_md: yes
  pdf_document: default
  rmarkdown::html_vignette:
    keep_md: yes
vignette: |
  %\VignetteIndexEntry{CSIDE-replicates} 
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup}
library(spacexr)
library(Matrix)
library(doParallel)
library(ggplot2)
datadir <- system.file("extdata",'SpatialRNA/Vignette',package = 'spacexr')# directory for sample Slide-seq dataset
datadir2 <- system.file("extdata",'SpatialRNA/Vignette_rep2',package = 'spacexr')
datadir3 <- system.file("extdata",'SpatialRNA/Vignette_rep3',package = 'spacexr')
datadir4 <- system.file("extdata",'SpatialRNA/Vignette_rep4',package = 'spacexr')
if(!dir.exists(datadir))
  dir.create(datadir)
savedir <- 'RCTD_results'
if(!dir.exists(savedir))
  dir.create(savedir)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  out.width = "100%"
)
```

## Introduction

Cell type-Specific Inference of Differential Expression, or CSIDE, is part of the *spacexr* R package for learning cell type-specific differential expression from spatial transcriptomics data. In this Vignette, we will demonstrate running RCTD and CSIDE on multiple replicates (slices) of cerebellum Slide-seq toy data. 

The ultimate goal will be to test for differential expression across two regions in the cerbellum. First, we will first use RCTD to assign cell types to a cerebellum Slide-seq dataset. We will define cell type profiles using an annotated single nucleus RNA-sequencing (snRNA-seq) cerebellum dataset. We will test for differential expression between the nodulus and anterior lobe regions.

## Data Preprocessing and Creating an RCTD.replicates object

First, we run RCTD on the data to annotated cell types (doublet mode). Please refer to the [spatial transcriptomics vignette](https://raw.githack.com/dmcable/spacexr/master/vignettes/spatial-transcriptomics.html) for more explanation on the RCTD algorithm, such as precise details for loading in the `SpatialRNA` object for this dataset. Since we need to load in the `SpatialRNA` object for four replicates, we will apply the function `read.SpatialRNA` to the directories of each of the four datasets to generate `puck1`, `puck2`, `puck3`, and `puck4`. Please note that in general, we recommend applying to each replicate a custom function that can load in a single replicate.

```{r load_spatial, message = FALSE}
### Load in/preprocess your data, this might vary based on your file type
puck1 <- read.SpatialRNA(datadir, count_file = "MappedDGEForR.csv", coords_file = 'BeadLocationsForR.csv')
puck2 <- read.SpatialRNA(datadir2, count_file = "MappedDGEForR.csv", coords_file = 'BeadLocationsForR.csv')
puck3 <- read.SpatialRNA(datadir3, count_file = "MappedDGEForR.csv", coords_file = 'BeadLocationsForR.csv')
puck4 <- read.SpatialRNA(datadir4, count_file = "MappedDGEForR.csv", coords_file = 'BeadLocationsForR.csv')
```

## Create explanatory variable / covariate

In order to run CSIDE, we will for each replicate create a explanatory variable (i.e. covariate) used for predicting differential expression in CSIDE. In general one should set the explanatory variable to biologically relevant predictors of gene expression such as spatial position. In this case, we will use the explanatory variable two divide each cerebellum replicate into two regions (left and right). In general, one can also create a custom function that will compute the explanatory variable on each `SpatialRNA` replicate. We will pass in a list of the explanatory variables to the `run.CSIDE.replicates` function.

Here, we also artifically upregulate the expression of the Kcnip4 gene in each dataset (in regions of high explanatory variable) to see whether CSIDE can detect this differentially expressed gene.

```{r SpatialRNA, results = 'hide', fig.height = 6, fig.width = 6}
explanatory.variable1 <- as.integer(puck1@coords$x > 3000)
names(explanatory.variable1) <- rownames(puck1@coords)
explanatory.variable2 <- as.integer(puck2@coords$x > 3000)
names(explanatory.variable2) <- rownames(puck2@coords)
explanatory.variable3 <- as.integer(puck3@coords$x > 3000)
names(explanatory.variable3) <- rownames(puck3@coords)
explanatory.variable4 <- as.integer(puck4@coords$x > 3000)
names(explanatory.variable4) <- rownames(puck4@coords)
exvar_list <- list(explanatory.variable1, explanatory.variable2, explanatory.variable3, explanatory.variable4)

# Differentially upregulate one gene
change_gene <- 'Kcnip4'
high_barc1 <- names(explanatory.variable1[explanatory.variable1 > 0.5])
puck1@counts[change_gene, high_barc1] <- puck1@counts[change_gene, high_barc1] * 2 
high_barc2 <- names(explanatory.variable2[explanatory.variable2 > 0.5])
puck2@counts[change_gene, high_barc2] <- puck1@counts[change_gene, high_barc2] * 2 
high_barc3 <- names(explanatory.variable3[explanatory.variable3 > 0.5])
puck3@counts[change_gene, high_barc3] <- puck3@counts[change_gene, high_barc3] * 2 
high_barc4 <- names(explanatory.variable4[explanatory.variable4 > 0.5])
puck4@counts[change_gene, high_barc4] <- puck4@counts[change_gene, high_barc4] * 2 
```

We also load in the `Reference` object, as follows.

```{r load_ref, warning = FALSE}
refdir <- system.file("extdata",'Reference/Vignette',package = 'spacexr') # directory for the reference
counts <- read.csv(file.path(refdir,"dge.csv")) # load in counts matrix
rownames(counts) <- counts[,1]; counts[,1] <- NULL # Move first column to rownames
meta_data <- read.csv(file.path(refdir,"meta_data.csv")) # load in meta_data (barcodes, clusters, and nUMI)
cell_types <- meta_data$cluster; names(cell_types) <- meta_data$barcode # create cell_types named list
cell_types <- as.factor(cell_types) # convert to factor data type
nUMI <- meta_data$nUMI; names(nUMI) <- meta_data$barcode # create nUMI named list
reference <- Reference(counts, cell_types, nUMI)
```
Now, we are in a position to create a `RCTD.replicates` object, an object that can store multiple `SpatialRNA` objects allowing for joint processing with RCTD and CSIDE. We pass in `spatial.replicates`, a list of the four `SpatialRNA` objects, `replicate_names`, a list of names for each replicate, and `group_ids`, a list of the group number for each replicate. In this case, all replicates will come from the same group. However, we have the option to assign different replicates to different groups, which will allow for extra variation to be modeled across groups.

```{r create, message = FALSE,warning = FALSE}
spatial.replicates <- list(puck1, puck2, puck3, puck4)
replicate_names <- c('1','2','3','4')
group_ids <- c(1,1,1,1)
myRCTD.reps <- create.RCTD.replicates(spatial.replicates, reference, replicate_names, group_ids = group_ids, max_cores = 2)
```

## Running RCTD

We are now ready to run RCTD with the `run.RCTD.replicates` function to assign cell types to these replicates. This function runs RCTD on each replicate in the `RCTD.replicate` object.

```{r run.RCTD, messages = FALSE}
myRCTD.reps <- run.RCTD.replicates(myRCTD.reps)
```
## Running CSIDE

After creating the explanatory variables and running RCTD, we are now ready to run CSIDE using the `run.CSIDE.replicates` function. Here, we pass in `exvar_list`, the list of explanatory variables for each replicate. We will use cell types 10 and 18 and a false discovery rate of 0.25. Next, we will set a gene threshold (i.e. minimum gene expression) of 0.01, and we will set a cell_type_threshold (minimum instances per cell type) of 5.

**Warning**: On this toy dataset, we have made several choices of parameters that are not recommended for regular use. On real datasets, we recommend first consulting the CSIDE default parameters. This includes `gene_threshold` (default 5e-5), `cell_type_threshold` (default 125), and `fdr` (default 0.01). Please see `?run.CSIDE.replicates` for more information on these parameters. 

```{r DEgenes}
#de
cell_types <- c('10','18')
myRCTD.reps <- run.CSIDE.replicates(myRCTD.reps, cell_types, exvar_list, gene_threshold = .01, 
                         cell_type_threshold = 5, fdr = 0.25) 

saveRDS(myRCTD.reps,file.path(savedir,'myRCTDde_reps.rds'))
```

## CSIDE results

After running CSIDE, it is time to examine CSIDE results. Note that the RCTD and CSIDE results on individual replicates are stored in `myRCTD.reps@RCTD.reps`, which is a list of `RCTD` objects containing individual CSIDE results as discussed in other vignettes. For this reason, we now focus on population-level CSIDE results across the four replicates.

We will focus on cell type 10. Furthermore, we will examine the original Kcnip4 gene, which was detected to be significantly differentially expressed in cell type 10. Other genes, including nonsignificant genes, can be observed in `myRCTD.reps@population_de_results`. As shown below, DE results on significant genes are stored in `population_sig_gene_df`. In particular, notice the columns `Z_est` (Z-score), `log_fc_est` (estimated DE coefficient), `sd_est` (standard error of DE coefficient), `p` (p-value), and `q_val` (q-value). Other columns of interest include `tau`, the estimated variance across samples, and `mean i` and `sd i`, which represent the DE point estimate and standard error on each of the four replicates.

```{r CSIDE_results,  fig.width = 8, fig.height=8}
#print results for cell type '18'
cell_type <- '10'
sig_gene <- change_gene
print(paste("following results hold for", sig_gene))
print(myRCTD.reps@population_sig_gene_df[[cell_type]]) 
```

Finally, we will save population CSIDE results in the `savedir` directory:

```{r results_save}
save.CSIDE.replicates(myRCTD.reps, file.path(savedir,'population'))
```

