---
output:  
  pdf_document: default
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, cache = T, results = 'hide')
```

# Finding Spatially Variable Genes

```{r libraries}
library(RCTD)
library(Matrix)
library(ggplot2)
library(ggpubr)
library(gridExtra)
library(reshape2)
library(Seurat)
```

### Setup and filter for genes expressed in CA3

```{r data}
#given a puck object, returns a puck with counts filtered based on UMI threshold and gene list
restrict_counts <- function(puck, gene_list, UMI_thresh = 1, UMI_max = 20000) {
  keep_loc = (puck@nUMI >= UMI_thresh) & (puck@nUMI <= UMI_max)
  puck@counts = puck@counts[gene_list,keep_loc]
  if(length(puck@cell_labels) > 0) #check cell_labels non null
    puck@cell_labels = puck@cell_labels[keep_loc]
  puck@nUMI = puck@nUMI[keep_loc]
  return(puck)
}

puck <- readRDS('../../Data/SpatialRNA/Puck_200115_08/puckCropped.RDS')
reference <- readRDS('../../Data/Reference/DropVizHC/scRefSubsampled1000.RDS')
cell_type_info <- get_cell_type_info(reference@assays$RNA@counts, reference@meta.data$liger_ident_coarse, reference@meta.data$nUMI)
gene_list <- intersect(rownames(cell_type_info[[1]]),rownames(puck@counts))
gene_list <- gene_list[rowSums(puck@counts[gene_list,]) >= 10]
gene_list <- gene_list[apply(cell_type_info[[1]][gene_list,],1,function(x) max(x)) >= 2e-5]
puck <- restrict_puck(puck, names(which(puck@nUMI >= 100)))
puck <- restrict_counts(puck, gene_list, UMI_max = 200000)
CA3_genes <- names(which(cell_type_info[[1]][gene_list,]$CA3*2 > apply(cell_type_info[[1]][gene_list,!(cell_type_info[[2]] %in% c('CA1','CA3','Denate','Neuron.Slc17a6', "Entorihinal", "Neurogenesis"))],1,max)))
CA3_genes <- CA3_genes[cell_type_info[[1]][CA3_genes,"CA3"] >= 2e-5]
```

### Compute expected cell type specific gene expression

```{r decompose}
#Command used to save the data from the gather_results.R script:
#save(puck_d, iv, results, file = 'Data/SpatialRNA/Puck_200115_08/results/final_analysis/gathered_results.RData')
#loading in that data:
load('../../Data/SpatialRNA/Puck_200115_08/results/final_analysis/gathered_results.RData')
puck_d <- get_decomposed_data(results$results_df, CA3_genes, puck, results$weights_doublet, cell_type_info)
```

### Find Spatially Variable Genes within CA3

```{r CA3_spatial}
cell_barc <- puck_d@cell_labels == "CA3" & puck_d@nUMI >= 300
gene_df <- puck_d@coords[cell_barc,]
gene_df$nUMI <- puck_d@nUMI[cell_barc]
get_int_genes <- function(gene_df,cell_type, gene_list_ct, puck_d, cell_barc) {
  pvals <- numeric(length(gene_list_ct)); names(pvals) = gene_list_ct
  cvs <- numeric(length(gene_list_ct)); names(cvs) = gene_list_ct
  toti = 0
  for(gene in gene_list_ct) {
    print(toti)
    toti = toti + 1
    gene_df$nUMI <- puck_d@nUMI[cell_barc]
    Tr = 100
    f_dist <- numeric(Tr)
    F_keep = 0
    big_count = 0
    for(i in 1:Tr) {
      gene_df$gene <- puck_d@counts[gene,cell_barc]/ puck_d@nUMI[cell_barc]
      if(i > 1)
        gene_df$gene <- sample(gene_df$gene)
      n = dim(gene_df)[1]
      fit <- loess(gene ~ x*y, span = 0.8, data = gene_df, degree = 1)
      #with(dat, plot(x, y, col = getcol(f), pch = 16, cex = 0.5, main = i))
      p <- fit$enp+1
      ss_total <-   sum((gene_df$gene- mean(gene_df$gene))^2)
      rss <- sum(fit$residuals^2)
      ss_reg <- ss_total - rss
      ms_reg <- ss_reg / (p-1)
      ms_res <- rss/(n-p)
      fstat <- ms_reg/ms_res
      if(i == 1) {
        gene_df$fitted <- fitted(fit)
        F_keep = fstat
        cvs[gene] <- sd(gene_df$fitted)/mean(gene_df$fitted)
        if(cvs[gene] < 0.5)
          break
      } else {
        if(fstat > F_keep)
          big_count = big_count + 1
      }
      if(big_count > 3)
        break
      #print(c(fstat, 1  - pf(fstat, p-1, n-p)))
      f_dist[i] <- fstat
    }
    pvals[gene] = (big_count + 1) / Tr
  }
  means_df <- data.frame(pvals,cvs)
  return(means_df)
}
means_df_CA3 <- get_int_genes(gene_df,"CA3", CA3_genes, puck_d, cell_barc)
int_genes_CA3 <- rownames(means_df_CA3[means_df_CA3$cvs >= 0.5 & means_df_CA3$pvals < 0.02,])
```

## Find Spatially Variable Genes Ignoring Cell Type

```{r global_spatial}
hippo <- CreateSeuratObject(puck@counts, project = "Slideseq",assay = "RNA",min.cells = 0,min.features = 0,names.field = 1,names.delim = "_",meta.data = NULL)
hippo <- NormalizeData(hippo, verbose = FALSE)
hippo <- FindVariableFeatures(hippo, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
hippo <- ScaleData(hippo)
hippo[['image']] <- new(Class='SlideSeq',assay='RNA',coordinates = puck@coords[colnames(puck@counts),])
resultsdir = "Data/Slideseq/NewCerPuck_190926_08/SeuratResults/"
DefaultAssay(hippo) <- "RNA"
hippo <- FindSpatiallyVariableFeatures(hippo, assay = "RNA", slot = "scale.data", features = VariableFeatures(hippo)[1:1000],
                                       selection.method = "moransi", x.cuts = 100, y.cuts = 100)
spatial_genes <- t(cell_type_info[[1]][head(SpatiallyVariableFeatures(hippo, selection.method = "moransi"),20),])
```

## Plot coefficient of variation of spatially variable genes, across and within cell types

```{r fig-1x, fig.height = 4, fig.width = 8, fig.align = 'center'}
CA3_spatial <- intersect(colnames(spatial_genes),rownames(means_df_CA3))
p_val <- means_df_CA3[int_genes_CA3,'pvals']
names(p_val) <- int_genes_CA3
write.csv(data.frame(p_val), file = "Results/CA3.csv")
all_cv <- apply(cell_type_info[[1]][gene_list,],1,function(x) sd(x)/mean(x))
plot_df <- data.frame(c(all_cv[colnames(spatial_genes)],sample(all_cv,50)), c(rep('glob',length(colnames(spatial_genes))),rep('null',50))) 
colnames(plot_df) <- c('quant','colr')
plot_df$cat <- 1
p2 <- ggplot(plot_df, aes(x = colr, y = quant)) +
  geom_jitter(alpha = 0.75) +
  geom_boxplot(fill = NA, width = 0.4, outlier.alpha = 0) +
  theme_classic() + scale_y_continuous(breaks = c(0,2,4), limits = c(0,4.2)) +xlab('Globally Variable Genes') + ylab('Coefficient of Variation Across Cell Types') +  scale_x_discrete(labels = c('Genes Detected Ignoring Cell Type','Randomly Selected Genes')) +  theme(axis.title.x=element_blank(),axis.text=element_text(size=8)) 

fc_df <- data.frame(means_df_CA3[int_genes_CA3,]$cvs, 'Local')
fc_df2 <- data.frame(means_df_CA3[CA3_spatial,]$cvs,'Global')
colnames(fc_df) <- c('score','class'); colnames(fc_df2) <- c('score','class')
fc_df <- rbind(fc_df2,fc_df)
p1 <- ggplot(fc_df, aes(x = class, y = score)) +
  geom_jitter(alpha = .75) +
  geom_boxplot(fill = NA, width = 0.4, outlier.alpha = 0) +
  theme_classic() + ylim(c(0,1.7)) + ylab('Coefficient of Variation Within CA3')  + scale_x_discrete(labels = c('Genes Detected Ignoring Cell Type','Genes Detected within Cell Type')) +  theme(axis.title.x=element_blank(),axis.text=element_text(size=7))


ggarrange(p2,p1,nrow = 1)
```

## Plot the gene Ptk2b in excitatory neurons and in general

```{r fig.height = 4, fig.width = 4, fig.align = 'center'}
my_mod <- function(p) {
  p + scale_x_continuous(breaks = c(1500,3500,5500), limits = c(1450,5700)) + scale_y_continuous(breaks =c(2000,3250,4500), limits = c(1800,4700)) + geom_segment(aes(x = 1700, y = 2100, xend = 2084.6, yend = 2100), color = "black")+ theme(axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(), axis.title.y=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank())+ theme(legend.position="top")
}
results_df <- results$results_df
gene = "Ptk2b"
my_class <- rep(0,length(colnames(puck@counts)))
names(my_class) <- colnames(puck@counts)
type_list <- c("CA1","CA3","Denate")
my_ind <- (results_df$spot_class != "reject" & results_df$first_type %in% type_list) | (results_df$spot_class == "doublet_certain" & results_df$second_type %in% type_list) 
my_ind2 <- (results_df$spot_class %in% c("singlet", "doublet_certain")) & !my_ind
my_class[names(which(puck@counts[gene,my_ind] < 0.1))] <- 1
my_class[names(which(puck@counts[gene,my_ind] >= 0.1))] <- 3
my_class[names(which(puck@counts[gene,my_ind2] < 0.1))] <- 2
my_class[names(which(puck@counts[gene,my_ind2] >= 0.1))] <- 4
my_barc <- names(my_class[my_class > 0])
p <- plot_class(puck, my_barc[order(my_class[my_barc])], factor(my_class)) 
p <- my_mod(p) + scale_color_manual(values=c("#CCE2EF","#F6DECC","#0072B2","#D55E00"))
ggarrange(p)
```

## 2D local regression smoothing of Rgs14 and Cpne9 genes within CA3 cell types

```{r fig-2fgddg, fig.height = 4, fig.width = 8, fig.align = 'center'}
my_pal = pals::brewer.ylorrd(20)[2:20]

my_mod <- function(p) {
  p + xlim(c(3800,5700)) + ylim(c(2100,3500)) + geom_segment(aes(x = 4000, y = 2400, xend = 4154, yend = 2400), color = "black")+ theme(axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(), axis.title.y=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank())+ theme(legend.position="top")
}

#int_genes <- c("Rgs14","Spink8",'Cpne9','Kcnf1')
int_genes <- c("Rgs14",'Cpne9')
MULT <- 500
p1 <- list()
p2 <- list()
for(ind in 1:length(int_genes)) {
  gene <- int_genes[ind]
  max_val <- 20*mean((puck_d@counts[gene,]/puck_d@nUMI)[cell_barc])
  cur_range <- c(0, MULT*.00087)
  if(ind == 2)
    cur_range <- c(0,MULT*.00025)
  barcodes <- rownames(gene_df)
  raw_df <- puck_d@coords[barcodes,]
  raw_df$val <- as.integer((puck_d@counts[gene,]/puck_d@nUMI)[barcodes] > 0.0002)
  raw_df <- raw_df[raw_df$val > 0,]
  raw_df$val <- factor(raw_df$val)

  
  
  gene_df$gene <- puck_d@counts[gene,cell_barc]/puck_d@nUMI[cell_barc]
  fit <- loess(gene ~ x*y, span = 0.8, data = gene_df, degree = 1)
  gene_df$fitted <- fitted(fit)
  plot_val <- pmax(0,gene_df$fitted)
  names(plot_val) <- as.character(which(cell_barc))
  min_val = MULT*min(plot_val);max_val = MULT*quantile(plot_val[gene_df$x  >= 3800 & gene_df$x < 5700 & gene_df$y > 2100 & gene_df$y < 3500],0.95)
  p2[[ind]]<- plot_puck_continuous(puck_d,barcodes,MULT*plot_val[names(puck_d@cell_labels)], ylimit = c(min_val,max_val), size = 1.25, alpha = 0.2)
  cur_range <- c(-.00001*MULT, signif(max_val,2))
  p2[[ind]] <- my_mod(p2[[ind]])+ ggplot2::scale_colour_gradientn(paste(gene, "Smoothed"), colors = pals::brewer.orrd(20)[2:20],limits = cur_range, breaks = c(0,signif(max_val,2)), labels = c(0,signif(max_val,2)))
  p2[[ind]] <- p2[[ind]] + geom_point(data=raw_df, aes(fill=val),,size = 0.1) + scale_fill_manual("Raw",values = c("#000000"),labels = c("")) + guides(fill = guide_legend(override.aes = list(size=2)))
}
#ggarrange(p1[[1]], p2[[1]], p1[[2]], p2[[2]],p1[[3]], p2[[3]],p1[[4]], p2[[4]], nrow = 4,ncol=2)
#ggarrange(p1[[1]], p2[[1]],p1[[2]], p2[[2]], nrow = 2,ncol=2)
ggarrange(p2[[1]], p2[[2]],nrow = 1,ncol=2)
```
