---
title: "RCTD and C-SIDE Documentation"
author: "Dylan Cable"
date: "12/22/2021"
output: github_document
---

```{r setup, include=FALSE}
library(spacexr)
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction

The *spacexr* manual can be found [here](https://github.com/dmcable/spacexr/tree/master/spacexr_manual_2.0.0.pdf).

Here, we will provide additional documentation on the *spacexr* R package. Other sources of spacexr documentation include the [vignettes](https://github.com/dmcable/spacexr/tree/master/vignettes) and the [Github issues](https://github.com/dmcable/spacexr/issues). Another great option for help on particular functions is to use the R manual files, which can be accessed with `?` or `help`. For example:

```{r cars}
?run.RCTD
?run.CSIDE.single
```

This page is intended to be a resource for frequently asked questions, tips and tricks, and odds and ends that are not addressed in other sources. Here, we will not be running code on real data as that is covered by the vignettes.

Terminology:

* *pixel*, synonymous with *spot* or *bead*, a measurement unit for spatial transcriptomics data that measures gene expression at a fixed location. We do not use the term *cell* because single *pixels* can sometimes contain mixtures of multiple cells.
* *explanatory variable*, synonymous with *covariate* or *predictive variable*, a vector with values at each pixel that can be used to explain gene expression.
* *design matrix*, the matrix containing all explanatory variables for explaining gene expression. Each explanatory variable appears as a single column in the matrix.

## Inputting data

### Creating a RCTD object

Both RCTD and C-SIDE require the construction of an `RCTD` object. In order to construct an RCTD object, you should use the `create.RCTD` function:

```{r}
myRCTD <- create.RCTD(spaceRNA, reference)
```

Here, `spaceRNA` represents the spatial transcriptomics dataset as a `SpatialRNA` object, described below. Also, `reference` is a `Reference` object (also described below) containing annotated cell types from a scRNA-seq dataset.

Important options for the `create.RCTD` function include (see `?create.RCTD` for details):

* `max_cores`: (default 4) number of cores used for running RCTD (and later C-SIDE). A rule of thumb is to use half the maximum number of cores on your machine.
* `MAX_MULTI_TYPES`: (default 4) if using RCTD multi mode, the maximum number of cell types per pixel.

In general, we will only highlight a subset of parameters/features, but curious users should always use the `help` function to find more information about the capabilities of a function.

We will now discuss how to create the `Reference` and `SpatialRNA` objects needed above.

### Creating a Reference object

The reference is created using the RCTD `Reference` constructor function:
```{r}
reference <- Reference(counts, cell_types, nUMI)
```

The function `Reference` requires 3 parameters: <br/>
1. `counts`: A matrix (or dgCmatrix) representing Digital Gene Expression (DGE). Rownames should be genes and colnames represent barcodes/cell names. Counts should be untransformed count-level data. <br/>
2. `cell_types`: 	
A named (by cell barcode) factor of cell type for each cell. The 'levels' of the factor would be the possible cell type identities. <br/>
3. `nUMI`: 	
Optional, a named (by cell barcode) list of total counts or UMI's appearing at each pixel. If not provided, nUMI will be assumed to be the total counts appearing on each pixel. <br/>

An option for the `Reference` function is `n_max_cells` (default 10000), which determines the maximum number of cells used from each cell type.

### Creating a SpatialRNA object

Similarly to the reference, first load the data into R, and then pass the data into the RCTD `SpatialRNA` constructor function.
```{r}
spaceRNA <- SpatialRNA(coords, counts, nUMI)
```
This function requires 3 parameters: <br/>
1. coords: A numeric data.frame (or matrix) representing the spatial pixel locations. rownames are barcodes/pixel names, and there should be two columns for 'x' and for 'y'.<br/>
2. counts: A matrix (or dgCmatrix) representing Digital Gene Expression (DGE). Rownames should be genes and colnames represent barcodes/pixel names. Counts should be untransformed count-level data. <br/>
3. nUMI: 	
Optional, a named (by pixel barcode) list of total counts or UMI's appearing at each pixel. If not provided, nUMI will be assumed to be the total counts appearing on each pixel. <br/>

## RCTD

### Running RCTD

We now show how to run RCTD to assign cell types.

```{r}
myRCTD <- run.RCTD(myRCTD, doublet_mode = 'doublet')
```

Here, we have three options for `doublet_mode`:
*`doublet mode`: at most 1-2 cell types per pixel, recommended for high spatial resolution technologies such as Slide-seq or MERFISH.
*`full mode`: no restrictions on number of cell types, recommended for low spatial resolution technologies such as Visium.
*`multi mode`: finitely many cell types per pixel, e.g. 3 or 4.

### RCTD results structure

The results of RCTD are located in the `@results` field. 

**Full mode results**: `@results$weights`, a data frame of cell type weights for each pixel for full mode. Weights should subsequently be normalized using the `normalized_weights <- normalize_weights(weights)` function. Weights will then add up to one for each pixel, representing cell type proportions.

**Doublet mode results:** The results of 'doublet_mode' are stored in `@results$results_df`. More specifically, the `results_df` object contains one column per pixel (barcodes as rownames). Important columns are:
* `spot_class`, a factor variable representing RCTD's classification in doublet mode: "singlet" (1 cell type on pixel), "doublet_certain" (2 cell types on pixel), "doublet_uncertain" (2 cell types on pixel, but only confident of 1), "reject" (no prediction given for pixel). Typically, `reject` pixels should not be used, and `doublet_uncertain` pixels can only be used for applications that do not require knowledge of all cell types on a pixel.
* Next, the `first_type` column gives the first cell type predicted on the bead (for all spot_class conditions except "reject"). 
* The `second_type column` gives the second cell type predicted on the bead for doublet spot_class conditions (not a confident prediction for "doublet_uncertain").

Additionally, `@results$weights_doublet` contains the doublet proportions for each cell type (`first_type` and `second_type`), respectively. One can use `get_doublet_weights` to convert the doublet mode weights to a matrix across all cell types.

**Multi mode results:** results consists of a list of results for each pixel, which contains `all_weights` (weights from full-mode), `cell_type_list` (cell types on multi-mode), `conf_list` (which cell types are confident on multi-mode) and `sub_weights` (proportions of cell types on multi-mode).

## C-SIDE

### Constructing design matrix / explanatory variables and running C-SIDE

In order to learn cell type-specific differential expression, one must first define one or multiple axis along which to expect differential gene expression. These axis are termed *explanatory variables*. Each explanatory variable is a vector of values, constrained between 0 and 1, with names matching the pixel names of the `myRCTD` object. For example, in the C-SIDE vignettes, many examples are provided of using e.g. spatial position, cell-to-cell interactions, or discrete regions as explanatory variables.

In the most common case, if a single explanatory variable is present, C-SIDE can be run using the `run.CSIDE.single` function:
```{r}
myRCTD <- run.CSIDE.single(myRCTD, explanatory.variable)
```

If multiple explanatory variables are present, then one can contruct the design matrix `X` directly. Several functions are provided in order to aid with design matrix construction including `build.designmatrix.single`, `build.designmatrix.regions`, and `build.designmatrix.nonparametric`. For example, for the case where one desires to test for differential expression across multiple (e.g. four) regions, one can construct `X` by organizing pixel barcodes into `region_list`, which is a list of character vectors, where each vector contains pixel names, or barcodes, for a single region:
```{r}
X <- build.designmatrix.regions(myRCTD, region_list)
```

After constructing the design matrix, one can use the general purpose `run.CSIDE` as follows:
```{r}
myRCTD <- run.CSIDE(myRCTD, X, barcodes, cell_types)
```

Here, `barcodes` represents a list of pixel names that should be included, while `cell_types` is a list of cell types that should be included in the model. 

In addition to using `run.CSIDE`, one can also consider `run.CSIDE.regions`, which can detect DE across multiple regions, and `run.CSIDE.nonparametric` for running nonparametric C-SIDE to fit smooth functions.

Consider also the following options for any of the `run.CSIDE.x` family of functions:
* `gene_threshold`: (default 5e-5) the minimum expression of genes to be included in the C-SIDE model.
* `doublet_mode`: (default TRUE) whether doublet mode or full mode weights are used in C-SIDE.
* `test_mode`: (default individual) if individual, tests DE coefficients individually for significant DE. If categorical (e.g. in case of multiple regions), then tests for differences across categories.
* `cell_type_threshold`: (default 125): minimum pixel occurrences of each cell type. We recommend that each cell type that is included occur at least on 50-100 pixels (see `count_cell_types` or `aggregate_cell_types` for counting cell type pixels).
* `cell_types_present`: List of other cell types that could be present in the dataset that could contaminate the measurements for cell types being analyzed.
* `params_to_test`: List of indices for which parameters should be tested.
* `test_genes_sig`: Whether C-SIDE should test genes for significance.
* `fdr`: False discovery rate for C-SIDE gene testing.

### C-SIDE results structure

C-SIDE results are stored in `myRCTD@de_results`. All gene expression results are computed in the natural log. The first object of interest containing C-SIDE results for all genes is `myRCTD@de_results$gene_fits`, referred to henceforth as `gene_fits`. This object is itself a list of several vectors and matrices of interest, including:

* `gene_fits$mean_val`: A genes by cell types matrix representing the C-SIDE point estimate of the differential expression coefficient of the explanatory variable. For example, if the explanatory variable represents two regions (0 and 1 variable), the `mean_val` represents the log-fold-change between regions. For example, `mean_val['geneA', 'Neuron'` would represent estimated DE for `geneA` in cell type `Neuron`. If multiple explanatory variables are present, one should consult the `all_vals` table instead, as described below.
* `gene_fits$all_vals`: All coefficients estimated by the C-SIDE model. Dimensions: genes X explanatory variables X cell types. For example, consider the case where the design matrix has categorical variables for four regions. In this case, `all_vals['geneA', 3, 'Neuron']` would represent the estimated coefficient (log gene expression) for geneA in cell type Neuron in region 3. 
* `con_mat`: A genes by cell types logical matrix representing convergence for each gene and for each cell type. For example, `con_mat['geneA', 'Neuron'` would represent convergence of the model on `geneA` in cell type `Neuron`. It is possible to converge for some cell types but not others, which occurs most commonly if one cell type is much more lowly expressed.
* `s_mat`: A genes by coefficients matrix representing the standard errors for each coefficient for each gene. We represent reshaping the matrix into a three dimensional tensor with `s_mat_3d <- get_standard_errors(s_mat)`. Dimensions: genes X explanatory variables X cell types. Then, e.g. `s_mat_3d['geneA', 3, 'Neuron']` would represent the coefficient standard error (log gene expression) for geneA in cell type Neuron and the third explanatory variable. 
* `sigma_g`: A vector, across genes, of the estimated standard deviation (magnitude) of pixel-specific overdispersion. A higher value of `sigma_g` indicates a higher level of noise compared to the Poisson distribution.

C-SIDE significant gene results are stored in `myRCTD@de_results$gene_fits`, which is a list of dataframes for each cell type. For example the following extracts the significant (as well as all genes including nonsignicant genes) gene dataframe for the 'Neuron' cell type:

```{r}
sig_df_neuron <- myRCTD@de_results$sig_gene_list[['Neuron']] # only significant genes
all_df_neuron <- myRCTD@de_results$all_gene_list[['Neuron']] # includes nonsignificant genes
```

This dataframe has rows for each significant gene. Columns will vary depending on C-SIDE mode, and one should consult the vignette of each mode for specific information. In most cases the following columns will be present:

* `log_fc`: the estimated differential expression, along the explanatory variable of interest, in natural log space.
* `se`: the standard error of this differential expression estimate.
* `paramindex_best`: the index of the coefficient that was tested for differential expression for this gene. This is relevant if multiple explanatory variables need to be distinguished.
* `Z_score`: the Z-score for a statistical Z-test, if applicable.
* `p_val`: the p-value for the statistical test of DE for this gene.
* `conv`: whether C-SIDE converged for this gene (required to achieve significance).

In the case of running C-SIDE with a single explanatory variable (`explanatory.variable`), the following columns will be present:

* `mean_0`: the estimated log gene expression of locations where `explanatory.variable = 0` (i.e. the intercept).
* `sd_0`: the standard error of `mean_0`.
* `mean_1`: the estimated log gene expression of locations where `explanatory.variable = 1` (i.e. the region with maximal explanatory variable).
* `sd_1`: the standard error of `mean_1`.

In the case of categorical-mode C-SIDE, C-SIDE performs pairwise Z-tests between pairs of regions, with the following columns will be present in these dataframes:

* `mean_i`: the estimated log gene expression in region i.
* `sd_i`: the standard error of this log gene expression estimate in region i.
* `sd_lfc`: the standard deviation of `mean_i`, across i.
* `log_fc_best`: the log-fold-change of the most differentially-expressed significant pair of regions.
* `sd_best`: the standard error of the log-fold-change estimate of this pair
* `p_val_best`: the p value of the differential expression for this pair.
* `paramindex1_best`: the parameter index for the first region in the most DE pair.
* `paramindex2_best`: the parameter index for the second region in the most DE pair.

If one would like to perform Z-tests on other pairs of regions, one can use the following formula to generate a Z-score (which can also be converted to a p-value using the standard Z-score method):

```
Z_i1_i2 <- (mean_i1 - mean_i2)/sqrt(sd_i1^2 + sd_i2^2)
```

### Multiple replicates

RCTD and C-SIDE can be run in batch mode on multiple replicates. This involves the creation of an `RCTD.replicates` object, which stores multiple `SpatialRNA` experimental replicates. Then, one can use `run.RCTD.replicates` and `run.CSIDE.replicates` to run RCTD and C-SIDE, respectively. This procedure is documented in detail in the [Population-level RCTD and C-SIDE vignette](https://raw.githack.com/dmcable/spacexr/master/vignettes/replicates.html).

## Frequently asked questions (FAQ) and Errors

* If I get an error with `spacexr`, should I post a Github issue? 

First of all, before posting an issue to Github, make sure to read the `spacexr` documentation (including this page and function documentation), Vignettes, and other Github issues. Next, there is an important distinction between caught and uncaught errors. If `spacexr` catches an error, then it will display a message such as:
```{r}
"Error in find_sig_genes_individual (cell_type, cell_types, gene_fits, gene_list_type, : "
"find_sig_genes_individual: cell type Endo has not converged on any genes. Consider removing this cell type from the model using the cell_types option."
```
In this case, `spacexr` has successfully detected the error and provides the user with feedback on how to fix the error. This is not a bug in `spacexr`. Rather, this is intended behavior for how to handle the error. In contrast, uncaught errors will trigger error messages outside the `spacexr` package. These are unintended error and should be reported, provided one is using the current package version and using the functions as intended.  

We outline several common errors and warnings below.

* C-SIDE error: cell type has not converged on any genes
```{r}
"Error in find_sig_genes_individual (cell_type, cell_types, gene_fits, gene_list_type, : "
"find_sig_genes_individual: cell type Endo has not converged on any genes. Consider removing this cell type from the model using the cell_types option."
```
This issue occurs when there are too few instances of a cell type. To debug this issue, one should first see how many of each cell type are included in the dataset by e.g. using the `aggregate_cell_types` function. Relatedly, the `cell_type_threshold` option to C-SIDE controls the minimum required number of each cell type. It is generally recommended that a cell type appear at least 100 times. This is especially the case for more complicated models with more parameters such as nonparametric mode. Nonparametric mode cannot be run without a few hundred instances per cell type. After counting cell types, cell types that are too rare should be removed from the list of cell types (controlled by the `cell_types` option to C-SIDE). Another potential cause of this issue is that there are too few genes in the `myRCTD` object. You can test the number of genes by looking at `rownames(myRCTD@originalSpatialRNA@counts)`. Having at least 1,000 genes will make it more likely that each cell type has at least a few genes for which counts are high.

* C-SIDE warning: C-SIDE vignette parameters
```{r}
"Warning message: In run.CSIDE.general(myRCTD, X1, X2, barcodes, cell_types, cell_type_threshold = cell_type_threshold, " 
"run.CSIDE.general: some parameters are set to the CSIDE vignette values, which are intended for testing but not proper execution. For more accurate results, consider using the default parameters to this function."
```
The C-SIDE vignettes have very small toy datasets and as such require abnormal parameter choices such as `cell_type_threshold` in order to run properly. These parameter values will not lead to good results on real data. The default C-SIDE parameter values would be a much better choice. This warning is to make users aware of this discrepancy.

